Estimating length of SARS-CoV-2 from doubly interval censored data with missed infections

### 2024-03-19 introduction.tex
proof intro
spelling / grammar check

All pandemics start as an \emph{outbreak}, a sudden and unexpected rise in the number of people with a disease in a local area; if this spreads over a larger geographic area (\eg a country) it becomes an \emph{epidemic}; if it spreads globally then it is a \emph{pandemic}~\autocite{grennanPandemic}.
The precise meaning of these terms is contested, determined implicitly by their usage rather than formal definitions~\autocite{morensPandemic,doshiElusive}.
I will refer to the SARS-CoV-2 epidemic within England (or the UK), and the global impact as the pandemic.

In this context, \emph{surveillance} is the ability to detect and monitor the spread of an outbreak.
Surveillance is important at every stage: early on, for detecting a new outbreak quickly and assessing its likely impact and later for monitoring the ongoing burden of the disease and the effectiveness of interventions~\autocite{whoFluSurveillance}.
In all cases, surveillance is required to inform public health policy.
In this thesis, I consider methods to improve surveillance, using the SARS-CoV-2 epidemic in England as a case study.

@@

Various forms of \emph{reproduction number} were prominent throughout the SARS-CoV-2 pandemic~\autocite{pellisEstimation}.
Broadly, a reproduction number describes the number of secondary infections generated by a single infection.
Here, following \textcite{pellisEstimation},  building on over a century of infectious disease dynamics literature, I define three that are broadly useful.

@@

The \emph{control reproduction number}, $\Rc$, is the mean number of secondary infections generated by a single infection with the behaviour at time $t$ but in a fully susceptible population.
$\Rc$ is also known as the \emph{reproduction number excluding immunity}.
It describes the effect on the epidemic of interventions (except vaccination) that have been put into place.
$\Rc$ is useful when comparing the effects of interventions at different points in time, when the immunity status of the population has changed.
For example, when comparing the English lockdown in March 2020 and January 2021, there are two effects in play: that there are more individuals with immunity in January 2021, and that the lockdown had different measures in place.
$\Rc$ focuses on the latter effect.

@@

Reproduction numbers are strongly connected to \emph{growth rates}.
Over short periods of time, epidemics tend to grow (or shrink) exponentially, and hence have an exponential growth rate, normally denoted $r$.
For a given disease and population, an analogous growth rate can be defined for each definition of the reproduction number~\autocite{pellisEstimation,paragGrowthRates,wallingaGI}.
The advantage of using the growth rate is that they are easier to estimate and capture how quickly spread happens in calendar time; however, they do not directly give the strength of intervention required to control the epidemic~\autocite{royalSocietyRnumber}.

@@

Growth rates and reproduction numbers only provide information on how an epidemic is changing.
Yet the size of the epidemic is just as, if not more, important~\autocite{pellisEstimation}.
\emph{Incidence} (the number of new infections in a given time period) is the primary metric for the magnitude of a pandemic.
Incidence informs the short-term future burden of the disease, for example, the future number of severe events (hospitalisations or deaths).
To translate incidence into the burden of the disease, a measure of severity is required; for example, the IFR (infection fatality rate), the proportion of infected individuals who die.
Estimation of severity is beyond the scope of this thesis.

\emph{Transmission} refers to the process by which an infectious disease generates new infections.
Estimating transmission refers to inferring the parameters of this process, including its components.
The components of transmission (\eg number of contacts per day and age-stratified rates) inform how interventions can be targeted.
They are also important for forecasting future incidence.

@@

The data sources used for outbreak surveillance are often observational in nature.
Therefore, they suffer from biases that can be hard to understand and quantify.
Here I discuss three types of data which are useful for surveillance.
Incidence-type and prevalence-type data can directly inform estimation of incidence and transmission, however, this estimation almost always relies on prior knowledge of epidemiological and/or biological quantities.
These quantities can be estimated from other studies, \eg cohort studies.

@@

Incidence-type data originate from observations of the consequences of a disease, \eg the development of symptoms or admission to hospital.
This type of data is typically affected by \emph{ascertainment bias}: the proportion of infections which are counted can vary over time or within different subgroups of the population~\autocite[chapter 9]{lashModern}.

@@

Prevalence-type data originate from observations of the number or proportion of a population in a specific disease state at a specific time.
For example, the number of hospital beds occupied with individuals with the disease or the proportion of the population with a marker of prior infection.
The issues associated with incidence-type data often also apply to prevalence-type data.

@@

Other epidemiological and/or biological quantities are normally required to estimate transmssion.
These depend on the model type, but could include the \emph{generation interval} (the time between an individual being infected and going on to infect further individuals) or duration of different disease stages.
Cohort studies, where a group of individuals are followed over time, are a study design which can estimate many of these quantities.

Well-designed cohort studies can provide direct estimates of incidence and/or prevalence in the cohort.
However, recruiting a representative sample of the population is challenging, and the results of these studies are often not generalisable to the whole population.
Furthermore, for these studies to be appropriately powered, they often require a large cohort, and hence are expensive to run.

@@

The first component required to estimate incidence and transmission is a model of the infection process, a mathematical description of the process through which incidence arises.
A spectrum of models exists, categorised by the strength of assumptions they make about the transmission process.
Stronger assumptions allow more insight into the transmission process and generally lower the variance of the estimates, but can introduce bias and are often computationally challenging to fit.

@@

Next, are semi-mechanistic models.
These have components of the transmission process in them but do not fully represent the process.
The most common type of semi-mechanistic model are those based on a \emph{renewal equation}.
The renewal equation~\autocite{fraserRenewal} for an infectious disease, in the most general form, is:
\begin{align}
    Z(t) &= \int_{0}^{t} \beta(t,\tau) Z(t-\tau) d \tau
\end{align}
where $Z(t)$ is the mean incidence at time $t$ and $\beta(t, \tau)$ is the instantaneous probability of an individual infected $\tau$ time units ago infects an individual at time $t$.
Therefore, the renewal equation relates incidence in the past to incidence in the future.
The effective reproduction number can be constructed as a function of $\beta(t, \tau)$ and the generation interval.
% A scaled measure of $I(t)$ (\eg hospital admissions or community test results) allows estimating $\beta$.
% The most common implementations are the R packages EpiNow2~\autocite{EpiNow2} and EpiEstim~\autocite{EpiEstim}.
Renewal equations do not feature further in this thesis; for details, including practical considerations when using them as the basis for inference, see \textcite{nashEstimating,gosticPractical}.

@@

\section{Disease models}

The \emph{disease model} links the transmission model to a population-level quantity which we will make observations of.
Examples are the incidence or prevalence of symptomatic individuals and admissions to hospital.

The choice of observation model can, in theory, be independent of the transmission model.
However, pragmatic considerations often mean that the observation model is chosen to be in a similar form to the transmission model.

Typically, the core of an observation model is a \emph{convolution equation} or an approximation of it.
This equation describes the expectation of the quantity at time $t$, $W_t$, conditional on the incidence prior to time $t$.
In discrete time, it is:
\begin{align}
    \E(W_t \mid Z_1, Z_2, \dots) &= \zeta \sum_{i=0}^\infty Z_{t-i} f(i)
    \label{intro:eq:disease-model}
\end{align}
where $Z_t$ is the incidence at time $t$; $\zeta$ is the proportion of infections that are ever observed; and $f(i)$ is the probability that an infection at time $t-i$ is observed at time $t$, conditional on it ever being observed.
$f$ is known as a \emph{delay distribution}.

Solving \cref{intro:eq:disease-model} for $Z$ is an ill-posed problem when $f(0)$ is small.
The forwards error of the solution is large when the ratio of the off-diagonal to diagonal elements is large~\autocite[chapter 8.2]{highamAccuracy}; for this situation, that is when the value of $\max_{i > 0} f(i) / f(0)$ is large.
A large forwards error means that the computed value is far from the true solution~\autocite[6--9]{highamAccuracy}.
Either an approach which gives a large $f(0)$ or smoothing on $Z_t$ is required.

@@

An observation model links $W_t$ to a possibly biased and/or noisy observation of it at time $t$.
Biases can arise in a variety of ways, \eg healthcare-seeking behaviour or tests which are less than 100\% accurate.
Bias can be modelled through additional parameters or data quantifying these issues.
Noise could come from a variety of sources, \eg sampling error, clustering, or unexplained overdispersion.
Noise is typically modelled by a parameteric distribution \eg the negative binomial distribution allowing for overdispersion compared to a Poisson process assumption.

@@

Two approaches are commonly used to estimate the duration of a disease stage: either modelling an underlying biomarker or modelling the duration directly~\autocite[e.g.][]{sweetingEstimating}.
Modelling a biomarker uses more information per observation, but requires stronger assumptions.

The biomarker approach models the trajectory of a biomarker over time in a cohort of individuals.
An individual is considered to have the disease while the biomarker is above or below some threshold.
Parametric assumptions are commonly required for this approach.
For example, a functional form of the behaviour over time of the biomarker is required.
In addition, if individual-level variation wants to be included, which is generally important, a distributional form of random effects is required.
This requires both sufficient observations per individual and an assumption on the form of the distribution.

Modelling the duration directly considers only the binary result at each test (positive or negative); for instance in a \emph{survival analysis} framework.
This approach generally requires fewer parameters, and no random effect(s), reducing the computational burden.
It is also easier to specify non-parametric forms of the distribution, which is important for SARS-CoV-2 as the tail behaviour may be important.
% A more detailed review of the literature is in the relevant chapters.
% Survival analysis is the area of statistics concerned with estimating the distribution of the times between two events.
% Other related approaches, such as a multi-state model (reviewed by \textcite{jacksonMSM}), could also be considered; multi-state models can be viewed as a generalisation of survival analysis.

In both cases, longitudinal data is required.
Therefore, cohort studies, where the same group of individuals are followed over time, are the most appropriate study design.

@@

The surveillance systems used in response to SARS-CoV-2 are historically unique.
Modern technology allowed large amounts of data to be collected, distributed, and analysed.
While the HIV/AIDS pandemic had inspired a large amount of statistical methodology to be developed, the pandemic was far slower moving.
Much analysis of SARS-CoV-2 data was done at great haste, requiring trade-offs to be made between timeliness and rigour of results presented.

Deep understanding of each data source's potential was not possible.
In this thesis, I will investigate the potential of the prevalence surveys in greater detail.

@@

Seroprevalence surveys were conducted throughout the epidemic, primarly in blood donors.
In addition to the normal problem of the delay between infection and antibody response, blood donors are a  convenience samples whose representativeness of the wider population is unclear.

The extraordinary impact of SARS-CoV-2 led the UK Government to fund two large prevalence surveys in representative samples of the general population.
These were not subject to the biases of community testing and observed the whole population, unlike severe events, although suffered from non-response bias.
Therefore, some modelling is required to borrow information across strata or time to provide stable estimates, as well as to correct for non-response bias (see \cref{E-intro:sec:cis}).
Despite this issue, they were considered the gold-standard metric.

@@

A second prevalence survey ran in the UK, known as REACT (REal-time Assessment of Community Transmission)~\autocite{rileyResurgence,rileyREACT}.
REACT was a series of cross-sectional surveys, run across 19 rounds between May 2020 and March 2022.
Each round consisted of at least 100,000 randomly-sampled individuals from the UK contributing swabs for testing and questionnaire data~\autocite{elliottTwin}.

@@

Many cohort studies to estiamte a wide variety of quantities ran throughout the pandemic.
In this thesis, I make use of two cohort studies with differing designs to estimate the duration or PCR positivity.
These designs have complementary strengths, and I combine information from the studies.
Both of the studies recruited individuals from the general population, aiming to observe a cohort of infections that are representative of the infections in the UK community.

% In addition to population-level surveillance, cohort studies help understand other aspects of the pathogen~\autocite{royalSocietyRnumber}.
% For example, contact tracing data can inform how likely it is for transmission to occur in different settings and/or between different groups of people.
The first study I use, ATACCC (Assessment of Transmission and Contagiousness of COVID-19 in Contacts)~\autocite{singanayagamCommunity,hakkiOnset}, performed daily testing of a small number of individuals with short follow-up.
The density of this sampling allows for reliable estimates of the viral load dynamics and other quantities of interest, most importantly the duration of detectability, but the results are limited because the small sample size and short follow-up means that the tails of these distributions cannot be estimated.
I describe ATACCC in detail in \cref{E-biology-data:sec:ataccc} and use it to estimate duration in \cref{E-ATACCC}.

The second study is the CIS (introduced in \cref{intro:sec:data-for-estimating}).
In addition to its role as a prevalence survey, the CIS followed the same individuals over a long period of time (over a year in many cases), collecting longitudinal data.
The large sample size and long follow-up mean that the tails of the distribution of the duration of positivity can be estimated.
However, the lack of frequent testing (at most weekly) means that the viral load dynamics cannot be estimated.
In particular, the vast majority of detected infections had only one or two positive tests, and hence the viral load trajectory of these infections cannot be estimated.
I describe the CIS in detail in \cref{E-intro:sec:cis} and use it to estimate duration in \cref{E-perf-test,E-imperf-test}.

@@

In the UK, the most modelling was produced by SPI-M-O (the Scientific Pandemic Influenza Group on Modelling, Operational sub-group).
SPI-M-O was a sub-committee of SAGE (the Scientific Advisory Group for Emergencies), generating evidence based on infectious disease modelling for policy development~\autocite{medleySPIM,govSPIMO}.
This used a wide range of  modelling approaches depending on the data available, which fluctuated throughout the epidemic, and the question being addressed.
Some were fit to data while others were purely used for simulation based on plausible parameter values.

Each week, the members of SPI-M-O would produce a set of estimates of the current state of the epidemic (\eg the value of $\Re$), known as \emph{nowcasting}.
These estimates, once any estimates unduly affected by issues (\eg data quality issues), were statistically combined into a consensus value~\autocite{parkCombining}.
The consensus value was published on the UK Government's website and informed its response to the epidemic~\autocite{govRnumber}.

@@

Some of these models were used to explore policy scenarios.
For example, each of the stages of lifting the lockdown in 2021 was supported by modelling studies informing the government's decision~\autocite{sageEvidence}.
These aided decision-makers in understanding plausible impacts of their actions.

@@

\emph{Backcalculation} is a modelling framework developed in the context of HIV/AIDS~\autocite{brookmeyerMethod}, but applied very broadly.
It specifies a particular combination of infection, disease, and observation model to relate times of HIV infection with times of AIDS diagnosis via the incubation period.
Further details are in the recent review \textcite{sunModeling}.
The theory behind the framework underpins much of the work in this thesis.

@@

To apply backcalculation to SARS-CoV-2 prevalence surveys, adaptations to the disease and observation models are needed (elaborated on in \cref{E-inc-prev}).
The disease model must consider that this is now prevalence-type, rather than incidence-type data.
This difference induces a dependency structure between the observations on each day.
The observation model must consider that this is now a survey, with sampling error, rather than assuming all infections are observed.

@@

\Textcite{hellewellPCRSensitivity} estimated the probability of an individual testing positive at each day after infection.
This measure is highly related to the duration of positivity.
The cohort in their study was drawn from healthcare workers, and hence was focused on working-age adults.

\Textcite{binnySensitivity} applied a slight variant of \textcite{hellewellPCRSensitivity}'s methodology to data from New Zealand.
The situation in New Zealand, who were successfully pursuing a zero-COVID strategy at this time, means that the individuals in the study were unusual.
For example, a third were travelers detected during their quarantine period on arrival to New Zealand, but would have had negative RT-PCR tests before their departure.

\Textcite{ealesCharacterising} estimated the duration of positivity from REACT data.
Specifically, individuals who tested positive in round 8 of REACT (6th--22nd Jan 2021) were invited to take two additional swab tests.
Defining the time of the initial positive test at time 0, \textcite{ealesCharacterising} estimated the probability of remaining positive at each day after the initial positive test.
They made a strong parametric assumption that the probability decreases exponentially, with a sensitivity analysis assuming a plateau before the decay.
This assumption is necessary because they had at most 17 days of follow-up, and hence could not estimate the tail of the distribution.
Overall, they estimate a median duration of 14.0 days (95\% CI: 12.9--15.4 days) but that this varied substantially by the viral load of an individual at the time of the initial positive test.
Individuals with a higher viral load were estimated to have a longer duration of positivity; the highest viral load individuals were estimated to have a median duration of 19.6 days (16.1--25.8 days).
This could be due to the higher viral load leading to a longer duration of positivity, or that individuals with a higher viral load were more likely to have been recently infected at the time of the initial positive test.
They cannot estimate how long before the initial positive test the individual was infected; therefore, they cannot distinguish between these two possibilities.
This also means they underestimate the duration of positivity.

For the purposes of this thesis, estimates applicable to the general population are required.
There are no reliable estimates of the full distribution of the duration of positivity in a general population cohort available in the literature.
Therefore, I estimate the duration of positivity from the ATACCC and CIS studies, as described above.

@@

Several previous studies have used SARS-CoV-2 prevalence surveys to estimating incidence and/or transmission.

\Textcite{abbottCISincidence} applied backcalculation to CIS data to estimate the incidence of SARS-CoV-2 in the UK.
They used a Gaussian process prior to smooth the incidence and stabilise the estimates.
\Textcite{abbottCISincidence} used \textcite{hellewellPCRSensitivity}'s estimates of the sensitivity of testing to inform the observation model, which are based on a longitudinal study in healthcare workers (see \cref{intro:sec:previous-duration-estimates}).
They fit to summary statistics of the CIS data, using a daily, independent Gaussian likelihood with parameters estimated from the posterior point estimate and 95\% credible interval published by the ONS.

\Textcite{colmanAscertainment} used CIS prevalence estimates to estimate the \emph{ascertainment rate} of community testing (the proportion of all infections that community testing detected).
Their methodology was similar to \textcite{abbottCISincidence}, including using estimates from \textcite{hellewellPCRSensitivity} for how the sensitivity of RT-PCR testing varies over an infection.
Rather than using a Gaussian process, as \textcite{abbottCISincidence} do, \textcite{colmanAscertainment} provide stability by assuming the ascertainment rate is assumed to be piecewise constant.
They do not fully propagate the uncertainty from the CIS prevalence estimates, considering only the bounds of the 95\% confidence intervals the ONS publishes for prevalence.

>>>>

All pandemics start as an \emph{outbreak}, a sudden and unexpected rise in the number of people with a disease in a local area; if this spreads over a larger geographic area (\eg a country) it becomes an \emph{epidemic}; if it spreads globally then it is a \emph{pandemic}~\autocite{grennanPandemic}.
The precise meaning of these terms varies, they are defined implicitly by their usage rather than formal definitions~\autocite{morensPandemic,doshiElusive}.
I will refer to the SARS-CoV-2 epidemic within England (or the UK), and the global impact as the pandemic.

In this context, \emph{surveillance} is the ability to detect and monitor the spread of an outbreak.
Surveillance is important at every stage: early on, for detecting a new outbreak quickly and assessing its likely impact and later for monitoring the ongoing burden of the disease and the effectiveness of interventions~\autocite{whoFluSurveillance}.
In all cases, surveillance informs public health policy.
In this thesis, I consider methods to improve surveillance, using the SARS-CoV-2 epidemic in England as a case study.

@@

Various forms of \emph{reproduction number} were prominent throughout the SARS-CoV-2 pandemic~\autocite{pellisEstimation}.
Broadly, a reproduction number describes the number of secondary infections generated by a single infection.
Here, following \textcite{pellisEstimation}, and building on over a century of infectious disease dynamics literature, I define three that are broadly useful.

@@

The \emph{control reproduction number}, $\Rc$, is the mean number of secondary infections generated by a single infection with the behaviour at time $t$ but in a fully susceptible population.
$\Rc$ is also known as the \emph{reproduction number excluding immunity}.
It describes the effect on the epidemic of interventions (except vaccination) that are in place at time $t$.
$\Rc$ is useful when comparing the effects of interventions at different points in time, when the immunity status of the population has changed.
For example, when comparing the English lockdown in March 2020 and January 2021, there are two effects in play: that there are more individuals with immunity in January 2021, and that the lockdown had different measures in place.
$\Rc$ focuses on the latter effect.

@@

Reproduction numbers are strongly connected to \emph{growth rates}.
Over short periods of time, epidemics tend to grow (or shrink) exponentially, and hence have an exponential growth rate, normally denoted $r$.
For a given disease and population, an analogous growth rate can be defined for each definition of the reproduction number~\autocite{pellisEstimation,paragGrowthRates,wallingaGI}.
The advantage of using the growth rate it is easier to estimate and captures how quickly spread happens in calendar time; however, they do not directly give the strength of intervention required to control the epidemic~\autocite{royalSocietyRnumber}.

@@

Growth rates and reproduction numbers only provide information on how an epidemic is changing.
Yet the size of the epidemic is just as, if not more, important~\autocite{pellisEstimation}.
\emph{Incidence} (the number of new infections in a given time period) is the primary metric for the magnitude of a pandemic.
Incidence informs the short-term future burden of the disease, for example, the future number of severe events (hospitalisations or deaths).
Incidence is translated into disease burden using a measure of severity is required; for example, the IFR (infection fatality rate), the proportion of infected individuals who die.
Estimation of severity is beyond the scope of this thesis.

\emph{Transmission} refers to the process by which an infectious disease generates new infections.
Estimating transmission refers to inferring the parameters of this process, including its components.
The components of transmission (\eg the number of contacts each individual makes per day) allow interventions to be targeted.
They are also important for forecasting future incidence.

@@

The data sources used for outbreak surveillance are often observational in nature.
Therefore, they suffer from biases that can be hard to understand and quantify.
Here I discuss three types of data which are useful for surveillance.
Incidence-type and prevalence-type data can directly inform estimation of incidence and transmission; however, this estimation almost always relies on prior knowledge of epidemiological and/or biological quantities.
These quantities can be estimated from other studies, \eg cohort studies.

@@

Incidence-type data originates from observations of the consequences of a disease, \eg the development of symptoms or admission to hospital.
This type of data is typically affected by \emph{ascertainment bias}: the proportion of infections which are counted can vary over time or within different subgroups of the population~\autocite[chapter 9]{lashModern}.

@@

Prevalence-type data originates from observations of the number or proportion of a population in a specific disease state at a specific time.
For example, the number of hospital beds occupied by individuals with the disease or the proportion of the population with a marker of prior infection.
The issues associated with incidence-type data often also apply to prevalence-type data.

@@

Other epidemiological and/or biological quantities are normally required to estimate transmission.
These depend on the model type but could include the \emph{generation interval} (the time between an individual being infected and going on to infect further individuals) or duration of different disease stages.
Cohort studies (following a group of individuals over time), are a study design which can estimate many of these quantities.

Well-designed cohort studies can provide direct estimates of incidence and/or prevalence in the cohort.
However, recruiting a representative sample of the population is challenging, and the results of these studies are often not generalisable to the whole population.
Furthermore, these studies often require a large cohort to be appropriately power.
Large cohort studies are expensive to run.

@@

The first component required to estimate incidence and transmission is a model of the infection process, a mathematical description of the process through which incidence arises.
A spectrum of models exists, categorised by the strength of assumptions they make about the transmission process.
Stronger assumptions allow more insight into the transmission process and generally lower the variance of the estimates but can introduce bias and are often computationally challenging to fit.

@@

Next, are semi-mechanistic models.
These have components of the transmission process in them but do not fully represent the process.
The most common type of semi-mechanistic model is those based on a \emph{renewal equation}.
The renewal equation~\autocite{fraserRenewal} for an infectious disease, in the most general form, is:
\begin{align}
    Z(t) &= \int_{0}^{t} \beta(t,\tau) Z(t-\tau) d \tau
\end{align}
where $Z(t)$ is the mean incidence at time $t$ and $\beta(t, \tau)$ is the instantaneous probability of an individual infected $\tau$ time units ago infects an individual at time $t$.
Therefore, the renewal equation relates incidence in the past to incidence in the future.
The effective reproduction number can be expressed as a function of $\beta(t, \tau)$.
% A scaled measure of $I(t)$ (\eg hospital admissions or community test results) allows estimating $\beta$.
% The most common implementations are the R packages EpiNow2~\autocite{EpiNow2} and EpiEstim~\autocite{EpiEstim}.
Renewal equations do not feature further in this thesis; for details, including practical considerations when using them as the basis for inference, see \textcite{nashEstimating,gosticPractical}.

@@

\section{Disease models} \label{intro:sec:disease-model}

The \emph{disease model} links the transmission model to a quantity, $W_t$, observable albeit possibly in a noisy and/or biased way (see next section).
Examples are the incidence or prevalence of symptomatic individuals and admissions to hospital.

The choice of observation model can, in theory, be independent of the transmission model.
However, pragmatic considerations often mean choosing the observation model to be in a similar form to the transmission model.

Typically, the core of an observation model is a \emph{convolution equation} or an approximation of it.
This equation describes the expectation of $W_t$, conditional on the incidence prior to time $t$.
In discrete time, it is:
\begin{align}
    \E(W_t \mid Z_1, Z_2, \dots) &= \zeta \sum_{i=0}^\infty Z_{t-i} f(i)
    \label{intro:eq:disease-model}
\end{align}
where $Z_t$ is the incidence at time $t$; $\zeta$ is the proportion of infections that ever appear in $W_t$; and $f(i)$ is the probability that an infection at time $t-i$ is included in count $W_t$, conditional on it ever being in one of the counts.
$f$ is known as a \emph{delay distribution}.

Solving \cref{intro:eq:disease-model} for $Z$ is an ill-posed problem when $f(0)$ is small.
The forwards error of the solution is large when the ratio of the off-diagonal to diagonal elements is large~\autocite[chapter 8.2]{highamAccuracy}; for this situation, which is when the value of $\max_{i > 0} f(i) / f(0)$ is large.
A large forwards error means that the computed value is far from the true solution~\autocite[6--9]{highamAccuracy}; this is avoided by either using an approach which gives a large $f(0)$ or smoothing $Z_t$.

@@

An observation model links $W_t$ to a possibly biased and/or noisy observation of it at time $t$.

Biases can arise in a variety of ways, \eg healthcare-seeking behaviour or tests which are less than 100\% accurate.
Bias can be modelled through additional parameters or data quantifying these issues.

Noise could come from a variety of sources, \eg sampling error, clustering, or unexplained overdispersion.
Noise is typically modelled by a parametric distribution \eg the negative binomial distribution allowing for overdispersion compared to a Poisson process assumption.

@@

Two approaches are commonly used to estimate the duration of a disease stage: either modelling an underlying biomarker or modelling the duration directly~\autocite[e.g.][]{sweetingEstimating}.
Modelling a biomarker uses more information per observation but requires stronger assumptions.

The biomarker approach models the trajectory of a biomarker over time in a cohort of individuals.
An individual is in the disease state while the biomarker is above or below some threshold.
This approach often requires parametric assumption(s).
For example, the functional form of the biomarker's trajectory.
In addition, if individual-level variation wants to be included, which is generally important, a distributional form of random effects is required.
This requires both sufficient observations per individual and an assumption on the form of the distribution.

Modelling the duration directly considers only the binary result at each test (positive or negative), for instance in a \emph{survival analysis} framework.
This approach generally requires fewer parameters, and no random effect(s), reducing the computational burden.
It is also easier to specify non-parametric forms of the distribution, which is important for SARS-CoV-2 as the tail behaviour may be important.
% A more detailed review of the literature is in the relevant chapters.
% Survival analysis is the area of statistics concerned with estimating the distribution of the times between two events.
% Other related approaches, such as a multi-state model (reviewed by \textcite{jacksonMSM}), could also be considered; multi-state models can be viewed as a generalisation of survival analysis.

Both cases require longitudinal data, suggesting that cohort studies are the most appropriate study design.

@@

The surveillance systems used in response to SARS-CoV-2 are historically unique.
Modern technology allowed large amounts of data to be collected, distributed, and analysed.
While the HIV/AIDS pandemic inspired the development of a large amount of statistical methodology, the pandemic was far slower moving.
Much analysis of SARS-CoV-2 data was done at great haste, requiring trade-offs to be made between timeliness and rigour of results presented.

A deep understanding of each data source's potential was not possible.
In this thesis, I will investigate the potential of the prevalence surveys in greater detail.
The disease model will be a prevalence model (defined further in \cref{inc-prev:sec:prevalence-process}) with the delay distribution being the duration of PCR positivity.

@@

Seroprevalence surveys were conducted throughout the epidemic, primarily in blood donors.
In addition to the normal problem of the delay between infection and antibody response, blood donors are a convenience sample whose representativeness of the wider population is unclear.

The extraordinary impact of SARS-CoV-2 led the UK Government to fund two large prevalence surveys in representative samples of the general population.
These were not subject to the biases of community testing and observed the whole population, unlike severe events, although suffered from non-response bias.
Therefore, some modelling is required to borrow information across strata or time to provide stable estimates, as well as to correct for non-response bias (see \cref{E-intro:sec:cis}).
Despite this issue, they are considered the gold-standard metric.

@@

A second prevalence survey ran in the UK, known as REACT (REal-time Assessment of Community Transmission)~\autocite{rileyResurgence,rileyREACT}.
REACT was a series of cross-sectional surveys, run across 19 rounds between May 2020 and March 2022.
Each round consisted of at least 100,000 randomly sampled individuals from the UK contributing swabs for testing and questionnaire data~\autocite{elliottTwin}.

@@

Many cohort studies to estimate a wide variety of quantities ran throughout the pandemic.
In this thesis, I make use of two cohort studies with differing designs to estimate the duration or PCR positivity.
These designs have complementary strengths, and I combine information from the studies.
Both of the studies recruited individuals from the general population, aiming to observe a cohort of infections that are representative of the infections in the UK community.

% In addition to population-level surveillance, cohort studies help understand other aspects of the pathogen~\autocite{royalSocietyRnumber}.
% For example, contact tracing data can inform how likely it is for transmission to occur in different settings and/or between different groups of people.
The first study I used, ATACCC (Assessment of Transmission and Contagiousness of COVID-19 in Contacts)~\autocite{singanayagamCommunity,hakkiOnset}, performed daily testing of a small number of individuals with short follow-up.
The density of this sampling allows for reliable estimates of the viral load dynamics and other quantities of interest, most importantly the duration of detectability, but the results are limited because the small sample size and short follow-up prohibit estimation of the tails of these distributions.
I describe ATACCC in detail in \cref{E-biology-data:sec:ataccc} and use it to estimate duration in \cref{E-ATACCC}.

The second study is the CIS (introduced in \cref{intro:sec:data-for-estimating}).
In addition to its role as a prevalence survey, the CIS followed the same individuals over a long period of time (over a year in many cases), collecting longitudinal data.
The large sample size and long follow-up allows estimation of the tails of the distribution of the duration of positivity.
However, the lack of frequent testing (at most weekly) prohibits estimation of the viral load dynamics.
In particular, the vast majority of detected infections had only one or two positive tests, and hence the viral load trajectory of these infections cannot be estimated.
I describe the CIS in detail in \cref{E-intro:sec:cis} and use it to estimate duration in \cref{E-perf-test,E-imperf-test}.

@@

In the UK, SPI-M-O (the Scientific Pandemic Influenza Group on Modelling, Operational sub-group) produced the most prominent modelling
SPI-M-O was a sub-committee of SAGE (the Scientific Advisory Group for Emergencies), generating evidence based on infectious disease modelling for policy development~\autocite{medleySPIM,govSPIMO}.
This used a wide range of  modelling approaches depending on the data available, which fluctuated throughout the epidemic, and the question being addressed.
Some were fit to data while others were purely used for simulation based on plausible parameter values.

Each week, the members of SPI-M-O would produce a set of estimates of the current state of the epidemic (\eg the value of $\Re$), known as \emph{nowcasting}.
The committee reviewed these estimates, removing any estimates unduly affected by issues (e.g.. data quality issues), and then formed a consensus value using a statistical combination of the estimates~\autocite{parkCombining}.
The consensus value was published on the UK Government's website and informed its response to the epidemic~\autocite{govRnumber}.

@@

Some of these models were used to explore policy scenarios.
For example, the decision at each of stage of lifting the lockdown in 2021 was informed by modelling studies~\autocite{sageEvidence}.
These aided decision-makers in understanding plausible impacts of their actions.

@@

\emph{Backcalculation} is a modelling framework developed in the context of HIV/AIDS~\autocite{brookmeyerMethod} but applied very broadly.
It specifies a particular combination of infection, disease, and observation model to relate times of HIV infection with times of AIDS diagnosis via the incubation period.
Further details are in the recent review \textcite{sunModeling}.
The theory behind the framework underpins much of the work in this thesis.

@@

The disease and observation models must be adapted to apply backcalculation to SARS-CoV-2 prevalence surveys (elaborated on in \cref{E-inc-prev}).
The disease model must consider that this is now prevalence-type, rather than incidence-type data.
This difference induces a dependency structure between the observations on each day.
The observation model must consider that this is now a survey, with sampling error.

@@

\Textcite{hellewellPCRSensitivity} estimated the probability of an individual testing positive each day after infection.
This measure is highly related to the duration of positivity.
The study's cohort consisted of healthcare workers, \ie working-age adults.

\Textcite{binnySensitivity} applied a slight variant of \textcite{hellewellPCRSensitivity}'s methodology to data from New Zealand.
The situation in New Zealand, who were successfully pursuing a zero-COVID strategy at this time, means that the individuals in the study were unusual.
For example, a third were travellers detected during their quarantine period on arrival to New Zealand but would have had negative RT-PCR tests before their departure.

\Textcite{ealesCharacterising} estimated the duration of positivity from REACT data.
Specifically, individuals who tested positive in round 8 of REACT (6th--22nd Jan 2021) were invited to take two additional swab tests.
Defining the time of the initial positive test at time 0, \textcite{ealesCharacterising} estimated the probability of remaining positive at each day after the initial positive test.
They made a strong parametric assumption that the probability decreases exponentially, with a sensitivity analysis assuming a plateau before the decay.
This assumption is necessary because they had at most 17 days of follow-up, and hence could not estimate the tail of the distribution.
Overall, they estimate a median duration of 14.0 days (95\% CI: 12.9--15.4 days) but with substantial positive correlation with each individual’s viral load at the time of the initial positive test (\ie a higher initial viral load was associated with a longer duration of positivity)
The individuals with the highest viral load individuals were estimated to have a median duration of 19.6 days (16.1--25.8 days).
The association could be due to individuals with a higher initial viral load being (on average) more recently infected at the time of the initial positive test.
The study design prohibits estimating how long before the initial positive test the infection occurred.
The same issue means the study underestimates the duration of positivity.

For the purposes of this thesis, I require estimates applicable to the general population.
There are no reliable estimates of the full distribution of the duration of positivity in a general population cohort available in the literature.
Therefore, I estimate the duration of positivity from the ATACCC and CIS studies.

@@

Several previous studies have used SARS-CoV-2 prevalence surveys to estimate incidence and/or transmission.

\Textcite{abbottCISincidence} applied backcalculation to CIS data to estimate the incidence of SARS-CoV-2 in the UK.
They used a Gaussian process prior to smooth the incidence and stabilise the estimates.
\Textcite{abbottCISincidence} used \textcite{hellewellPCRSensitivity}'s estimates of the sensitivity of testing to inform the observation model, which are based on a longitudinal study in healthcare workers (see \cref{intro:sec:previous-duration-estimates}).
They fit to summary statistics of the CIS data, using a daily, independent normal likelihood with parameters estimated from the posterior point estimate and 95\% credible interval published by the ONS.

\Textcite{colmanAscertainment} used CIS prevalence estimates to estimate the \emph{ascertainment rate} of community testing (the proportion of all infections that community testing detected).
Their methodology was similar to \textcite{abbottCISincidence}, including using estimates from \textcite{hellewellPCRSensitivity} for how the sensitivity of RT-PCR testing varies over an infection.
Rather than using a normal process to provide stability, as \textcite{abbottCISincidence} do, \textcite{colmanAscertainment} assume the ascertainment rate is piecewise constant.
They do not fully propagate the uncertainty from the CIS prevalence estimates, considering only the bounds of the 95\% confidence intervals the ONS publishes for prevalence.
